A Standard for the Scholarly Citation of Archaeological Data

### 2017-09-12 analysis/paper/paper.Rmd
writing about ethics

---
title: "A Standard for the Scholarly Citation of Archaeological Data"
author:
  - "Ben Marwick"
  - "Suzanne E. Pilaar Birch"
  - 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

In the pilot studies reported above we see three common approaches to sharing data: sending files privately by email, following a personal request; including data as supplementary files submitted for publication with a journal article; and sharing data by depositing files in a trustworthy, public, DOI-issuing repository. The third method, via a public repository, is by far the most widely recommended method both specifically by archaeologists (Kintigh 2006; Kansa et al. 2011; Kansa and Kansa 2013), and in other sciences (cf. Cranston et al. 2014; Penev et al. 2017; Mounce 2014; Thessen & Patterson 2011; Whitlock 2011).	There are many guidelines and conventions from other disciplines that are relevant to sharing archaeological data (ICPSR 2012; Strasser et al. 2012). Table XXX summarises some of these, and recommendations specifically written for archaeologists can be found in UK Archaeology Data Service (n.d.), ARCHES project. (n.d.) [archaeology data service manuals]. Many of these guidelines may create new technical and practical challenges to archaeologists who are not accustomed to sharing data, and many not want to dedicate resources to sharing data. However, we should not let the perfect be the enemy of the good. In our own work we try to follow these guidelines with no dedicated resources for data sharing, and with an incomplete knowledge of some of the technical details of repository operations. While we recognise that our data sharing pracitses fall short of all the guidelines, we still see value in sharing data imperfectly, rather than not at all.

@@

[ ] - datacite method
http://journals.plos.org/plosbiology/article?id=10.1371/journal.pbio.2001414

<!---
-In-text citation:

 “This paper uses data from the [name] data repository at https://doi.org/***** (Jones et al. 2008a), first described in Jones et al. 2008b. “

 Data reference and article reference in reference list:

 Jones A, Bloggs B, Smith C (2008a). <Title of data package>. <Repository name>. doi: https://doi.org/#####. [Version and/or date of access].
 -->

 # Ethical limitations to data sharing

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

---
title: "A Standard for the Scholarly Citation of Archaeological Data"
author:
  - "Ben Marwick"
  - "Suzanne E. Pilaar Birch"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

In the pilot studies reported above we see three common approaches to sharing data: sending files privately by email, following a personal request; including data as supplementary files submitted for publication with a journal article; and sharing data by depositing files in a trustworthy, public, DOI-issuing repository. The third method, via a public repository, is by far the most widely recommended method both specifically by archaeologists (Kintigh 2006; Kansa et al. 2011; Kansa and Kansa 2013), and in other sciences (cf. Cranston et al. 2014; Penev et al. 2017; Mounce 2014; Thessen & Patterson 2011; Whitlock 2011).	There are many guidelines and conventions from other disciplines that are relevant to sharing archaeological data (ICPSR 2012; Strasser et al. 2012; Wilkinson et al. 2016). Table XXX summarises some of these, and recommendations specifically written for archaeologists can be found in UK Archaeology Data Service (n.d.), ARCHES project. (n.d.) [archaeology data service manuals]. Many of these guidelines may create new technical and practical challenges to archaeologists who are not accustomed to sharing data, and many not want to dedicate resources to sharing data. However, we should not let the perfect be the enemy of the good. In our own work we try to follow these guidelines with no dedicated resources for data sharing, and with an incomplete knowledge of some of the technical details of repository operations. While we recognise that our data sharing pracitses fall short of all the guidelines, we still see value in sharing data imperfectly, rather than not at all.

@@

The Joint Declaration of Data Citation Principles is the current standard in may disciplines for data citation within any research object (Data Citation Synthesis Group 2014). The result of these principles is a citation formation as follows (the exact order of the items may vary slightly across different referencing styles):

Creator (PublicationYear). Title. Version. Publisher. Identifier

The 'PublicationYear', 'Title' and 'Publisher' are familiar from traditional citation formats. 'Creator' is equivalent to the author or primary collector and arranger of the data. The 'Version' and 'Identifier' fields are unique to data citation, and can take a variety of forms. The identifier is a special type of persistent internet address. The identifier must have two important qualities: the address contains some metadata (such as the identity of the publisher), and they remained fixed to a specific file or bundle of files over the lifetime of those objects, even if their location on the internet changes. There are several varieties of these in use, with DOIs (digital object identified) being the most common form. When a data publisher changes the internet address of a data set, the are responsible for informing the DOI-issuing authority so that the DOI resolves to the new location of the files.

A DOI consists of three parts:

- the proxy, usually https://doi.org but sometimes also https://hdl.handle.net/ or just doi:
- the prefix, usually a string of numbers that is unique to the organisation publishing the DOI. For example, tDAR prefix is 10.6067
- the suffix, a random-looking string of any combination of characters and numbers that uniquely identified the file or bundle of files. This is also known as a  local identifier (local ID) or an 'accession', an identifier which, by itself, is only guaranteed to be locally unique within the database or source (McMurry et al. 2017)

A DOI alone is not suffcient for citing a dataset because, like many things onthe internet, the files can change over time. For example we might upload data files to the repository when we submit a paper for peer review, then get a DOI and cite that in our manuscript so the peer reviewers can see them (some services provide a private URL the can be shared with reviewers but it not publicly accessible or discoverable by a search engine). Then when we resubmit our manuscript after peer review, we might update the repository to include additional data files relating to additional analyses requested by the reviewers. We still have the same DOI in our manuscript for this data set, but the files have changed because of the revisions we did in response to the peer reviewers' comments. Most data repositories will keep each version of the files that you upload, so in this scenario there will be at least two versions of the files in our data repository. Even if we do not follow this exact sequence of updating the files, we may still change them in the future, after our paper has been published. Because of this potential to change the files stored at the DOI, it is necessary to identify specific file versions in our citation of the data. A simple method to record the version is to include the date that the data files were accessed, as when citing any other resource found on the internet. Some repositories also provide DOIs that have a version number suffix like '.v2', '.2', '/2'. However this is not formally a functionality of the DOI system, so including the date accessed is a more robust and general method of indicating the version.

When citing data in the text of a manuscript or report, the ideal practice is to cite the dataset and the traditional publication that first described it, like so (Penev et al. 2017):

"This paper uses data from the [name] data repository at http://doi.org/xxx/xxx (Author YYYY), first described in Author YYYY"

For example:

"This paper uses data from the Open Science Framework data repository at http://doi.org/10.17605/OSF.IO/RTZTH (Marwick 2017), first described in Marwick et al. 2017."

Citing both the dataset and the publication helps to resolve ambiguity about the source of the data, and gives context to the data. Furthermore the authorship of the data files may differ from those of the author's paper describing the data, and it is important that both authors (or sets of authors) recieve credit for their work through the normal mechanisms of citation recognition.

When referring to newly published data in the paper in which these data are first described, the citation statement should be included in the body of the paper, in the Material and Methods section. In addition, the formal data reference should be included in the paper's reference list, using the recommended journal's reference format. For example:

“The data underpinning the analysis reported in this paper were deposited in the Open Science Framework Repository at http://dx.doi.org/10.17605/OSF.IO/RTZTH (Marwick 2017).

With the reference list of that paper including Marwick (2017) as:

Marwick, B. (2017). Dataset and R code for "Movement of lithics by trampling: An experiment in the Madjedbebe sediments", http://doi.org/10.17605/OSF.IO/RTZTH, Accessed 7 Sept 2017

# Ethics and data sharing

A full treatment of the ethical issues of data sharing is beyond the scope of this paper, but we can highlight some points specifically relevant to the question of when archaeological data. Much of the discussion on ethics and data sharing generally relates to appropriate attribution (assigning credit for the data) in data reuse (Duke et al. 2013). The default position for archaeological data, like data from many other sciences, is that they are part of the global human patrimony, and should be accessible to other scholars and the public (Zimmerman et al. 2003; Vitelli & Colwell-Chanthaphonh 2006). However, archaeologists, like other researchers working with indigenous communities and governments, generate research data that is intellectual property with many stakeholders. With diverse stakeholders, archaeological data can become entangled in economics, nationalism, cultural politics, and identity (Nicholas and Wylie 2009). Often, one of these stakeholders is an indigenous community, and archaeologists may find differences between Western and indigenous conceptions of what data are, how they may be generated, used, shared, and 'owned' (Harding et al. 2012).

One of these differences is that Western researchers often have individual autonomy in sharing data, and once data are shared, it is free for all to use. In contrast, within indigenous communities, data are part of the group's identity and property, akin to tangible hertiage (Nicholas 2014), with certain members entrusted to keep that data on behalf of the group, and sometimes it may be inappropriate for these people to share the data with other members of the group, or beyond the group (Tsosie 2007). Differences between Western and indigenous concepts or data are not only cultural - substantial inequalities exist in access to resources for converting archaeological data into economic and political benefit (e.g. graduate degrees, employment, publications, grants, promotions). On one hand, making archaeological data openly available benefits indigenous communities, because it removes some of the financial barriers to accessing that data (e.g. subscription fees to journals). On the other hand, this may be little more than a useless courtesy (Hymes 1972), because simply making data openly available does not give indigenous people access to the contexts where the data can be used to generate economic and political benefits. Thus, open data has different values for Western and indigenous communities, and if archaeologists want to make a compelling case to their indigenous partners for data sharing, they should also make a committement to reducing some of the inequalities in making use of the data, for example, by providing some training in data analysis, or adapting the data into a context that is valued by the indigenous community.

In short, there is more to the ethics of data sharing than avoiding risk of damage. Redacting the locations of archaeological sites to prevent looting as an obvious case of preparing data for sharing to reduce the risk of damage to the sites. However, because of cultural, economic and political differences between Western archaeologists and the communities they work with, an archaeologist cannot consider their own risk assessment to be comprehensive (Nicholas and Wylie 2012). It is crucial for archaeologists to discuss data sharing plans during negotiatations with representatives authorized by the indigenous peoples whose cultural heritage is the subject of investigation. This negotiation process is important because only indigenous peoples themselves can identify potential adverse outcomes to data sharing, and they can do this only if they understand the proposed research and anticipated results. The concrete outcome of this process is often a Memorandum of Understanding between the indigenous community and the archaeologist's employer that documents the ourcome of negotiations about data sharing. If the archaeologist's indigenous partners have reasons not to share data in public, then the archaeologist should respect that position, and note in publication that the data are restricted at the request of the indigenous collaborators.

@@

Duke, C. S., & Porter, J. H. (2013). The Ethics of Data Sharing and Reuse in Biology. BioScience, 63(6), 483–489. https://doi.org/10.1525/bio.2013.63.6.10

@@

Hymes, D. (1972). 1974. Reinventing anthropology.

@@

McMurry, J. A., Juty, N., Blomberg, N., Burdett, T., Conlin, T., Conte, N., … Parkinson, H. (2017). Identifiers for the 21st century: How to design, provision, and reuse persistent identifiers to maximize utility and impact of life science data. PLOS Biology, 15(6), e2001414. https://doi.org/10.1371/journal.pbio.2001414

@@

Nicholas, G. (2014). Indigenous cultural heritage in the age of technological reproducibility: Towards a postcolonial ethic of the public domain. Dynamic Fair Dealing: Creating Canadian Culture Online, 213-24.

@@

Nicholas, G. P., & Wylie, A. (2012). Do Not Do Unto Others…’: Cultural misrecognition and the harms of appropriation in an open-source world. Appropriating the Past: Philosophical Perspectives on Archaeological Practice, 195-221.

Nicholas, G. P., & Wylie, A. (2009). Archaeological finds: legacies of appropriation, modes of response. The Ethics of Cultural Appropriation, 11-54.

@@

Penev L, Mietchen D, Chavan V, Hagedorn G, Smith V, Shotton D, Ó Tuama É, Senderov V, Georgiev T, Stoev P, Groom Q, Remsen D, Edmunds S (2017) Strategies and guidelines for scholarly publishing of biodiversity data. Research Ideas and Outcomes 3: e12431. https://doi.org/10.3897/rio.3.e12431

@@

Tsosie, R. (2007). Cultural challenges to biotechnology: Native American genetic resources and the concept of cultural harm. The Journal of Law, Medicine & Ethics: A Journal of the American Society of Law, Medicine & Ethics, 35(3), 396–411. https://doi.org/10.1111/j.1748-720X.2007.00163.x

@@

Vitelli, K. D., & Colwell-Chanthaphonh, J. S. (Eds.). (2006). Archaeological ethics. Rowman Altamira.

@@

Wilkinson, M. D., Dumontier, M., Aalbersberg, Ij. J., Appleton, G., Axton, M., Baak, A., Bourne, P. E. (2016). The FAIR Guiding Principles for scientific data management and stewardship. Scientific Data, 3, 160018. https://doi.org/10.1038/sdata.2016.18

@@

Zimmerman, L. J., Vitelli, K. D., & Hollowell-Zimmer, J. (Eds.). (2003). Ethical issues in archaeology. Rowman Altamira.
